---
title: "ice_cover"
format: html
editor: visual
---

```{r, message = false}
library(tidyverse)
library(here)
library(ggbeeswarm)
library(patchwork)
```

-   Read in both files (ice cover duration and meteorological data). Note that the ice cover data has some NA values stored as -999.

```{r, message= FALSE, warning=FALSE}
ice_cover <- read_csv(here::here("data", "ntl33_v7.csv"), na = "-999")
meteorological <- read_csv(here::here("data", "ntl20_v6.csv"), na = "-999")
```

-   Create a visualization that compares ice duration across different lakes in the Madison Lake Area. Include both the actual ice duration values *and* some representation of summary statistics for each lake. Careful if you use a boxplot over a jitter or beeswarm - what values might be doubly represented? Can you find an argument that would help you remove those from the boxplot?

```{r}

ice_cover_duration <- ice_cover |> 
  ggplot(aes(x = lakeid, y= ice_duration, na.rm = TRUE)) +
  geom_beeswarm(size = 0.5) + geom_boxplot(fill = NA, outlier.color = NA)
ice_cover_duration
```

-   Find the mean ice cover duration in Madison Lake Area by year (i.e., you'll only be grouping by the year variable here). Understanding check: what are you finding the mean of here? Why are there multiple observations for each year?)

    ```{r}
    mean_ice_cover <- ice_cover |> 
      group_by(year4) |> 
      summarize(mean_ice_duration = mean(ice_duration, na.rm = TRUE))

    mean_ice_cover
    ```

-   Create an exploratory visualization of mean ice cover duration by year for the Madison Lake Area. Add a sentence or two below this exploratory graph describing the overall trend(s) you observe.

    ```{r}
    ggplot(mean_ice_cover, aes(x = year4, y = mean_ice_duration, color = mean_ice_duration)) +
      geom_smooth()
    ```

-   Find the mean air temperature (using the daily average air temperature - adjusted) in Madison Lake Area **by year** using **only observations from winter months (December, January, February)**

-   Create an exploratory visualization of mean winter temperatures for the Madison Lake Area. Add a sentence or two below this exploratory graph describing the overall trend(s) you observe.

-   Join the mean winter air temperatures to the mean ice cover duration data you found above

-   Create an exploratory scatterplot of mean winter temperatures versus mean ice duration. Add a sentence or two below this exploratory graph describing the overall trend(s) you observe.

-   Explore thaw dates since 1970. Create a subset of the ice cover duration data since 1970 (ok to include 1970), then convert the `ice_off` column to a `Date`. Use `lubridate::yday()` to pull the numeric *day of the year* from that Date. Create an exploratory plot of the ice off day (numeric day-of-the-year) from 1970 - present. Add a sentence or two below this exploratory graph describing the overall trend(s) you observe

```{r, message= FALSE, warning=FALSE}
icecover <- read_csv(here::here("data", "ntl33_v7.csv"), na = "-999")
weather <- read_csv(here::here("data", "ntl20_v6.csv"), na = "-999")
```

```{r}

ggplot(data = icecover, aes(x = lakeid, y = ice_duration)) +
  geom_jitter(width = 0.2, color = "gray50") +
  geom_boxplot(width = 0.3, fill = NA, size = 1, outlier.color = NA)

mean_duration <- icecover %>% 
  group_by(year4) %>% 
  summarize(mean_duration = mean(ice_duration, na.rm = TRUE))

ggplot(data = mean_duration, aes(x = year4, y = mean_duration)) +
  geom_line()
weather_summary <- meteorological %>% 
  filter(month %in% c(12, 1, 2)) %>% 
  group_by(year4) %>% 
  summarize(mean_temp = mean(ave_air_temp_adjusted, na.rm = TRUE))

ggplot(data = weather_summary, aes(x = year4, y = mean_temp)) +
  geom_line()

ice_weather_join <- mean_ice_cover %>% 
  full_join(weather_summary)

ggplot(data = ice_weather_join, aes(x = mean_temp, y = mean_duration, na.rm = TRUE)) + geom_point()

thaw_dates <- icecover %>% 
  filter(year4 >= 1970) %>% 
  mutate(ice_off = lubridate::mdy(ice_off)) %>% 
  mutate(day_ice_off = lubridate::yday(ice_off))

ggplot(data = thaw_dates, aes(x = year4, y = day_ice_off)) +
  geom_point()
```

